### 2014-01-16 thesis-proposal/problem.tex
Full rough draft.

% !TEX root = omar-thesis-proposal.tex
\vspace{-25pt}
\section{Motivation}\label{motivation}
Specifying and implementing a programming language together with its supporting tools (collectively, a \emph{programming system}) that has sound theoretical foundations, helps users identify and fix errors as early as possible, supports a natural programming style, and performs well across diverse problem domains and hardware platforms remains a grand challenge in computing. Due to the increasing diversity and complexity of modern problem domains and hardware platforms, it has become clear that no small  collection of general-purpose constructs backed by a simple, all-purpose implementation can fully satisfy these criteria in all cases. Instead,  researchers and domain experts (collectively, \emph{providers}) continue to develop  specialized notations,  static semantics, dynamic semantics, implementation strategies, optimizations, run-time systems and tools (collectively, \emph{features})  designed to address these modern challenges in different circumstances. %, researchers and domain experts must continue to design and develop, in a principled manner, specialized notations, type systems, optimizations, run-time systems and tools (collectively, \emph{features}) to support developers as they work in increasingly complex problem domains and on modern hardware systems.

@@

Ideally, a provider would develop and distribute such new features orthogonally, as libraries, so that client developers could granularly choose those that best satisfy their needs. Unfortunately this is often infeasible because libraries are vehicles for specifying a program in terms of a monolithic collection of existing features. From the perspective of a library, the language's syntax, type system and semantics are fixed in advance, the compiler and run-time system are ``black box'' implementations of this fixed language, and the other tools, like code editors and debuggers, operate according to domain-agnostic protocols also based only on a fixed language specification. As a result, providers of new system features must often take \emph{language-external approaches}, sometimes designing a new system altogether. We will argue that such approaches are problematic, and that taking them has led to an unnecessary gap between research and practice. In place of these approaches, we will advocate for \emph{language-integrated extensibility mechanisms}, and show how, by organizing new features around {types} enforcing certain constraints, these mechanisms can be made both safe and highly expressive. We call a type that introduces new features an \emph{active type} and programming systems organized around active types \emph{actively-typed programming systems}.
%But taking a \emph{language-internal approach} to implementing a feature is the most practical. If a feature can be realized by creatively using existing language constructs and distributed as a library, clients face fewer barriers to adoption because it is easy to integrate library-based features into existing projects gradually and granularly and they leverage well-understood and well-developed mechanisms.
% But taking this approach is often \emph{not} possible today %We call designs \emph{monolithic programming systems}.

@@

\begin{enumerate}
\item \textbf{Built-in syntax for pattern literals} (e.g. \cite{perlre}) so that malformed patterns result in intelligible \emph{compile-time} parsing errors. In languages lacking such literals, run-time errors relating to pattern syntax can occur, even in well-tested, deployed code \cite{spishak2012type}. 
\item \textbf{Typechecking logic} that ensures that key constraints related to regular expressions are not violated at compile-time:
	\begin{enumerate}
	\item only appropriate values are spliced into regular expressions, to avoid splicing errors and injection attacks \cite{owasp2013}
	\item out-of-bounds backreferences are not used \cite{spishak2012type}
	\item strings known to be in the language of a regular expression are given a type that tracks this information and ensures that string manipulation operations do not inadvertently lead to a string that is not in the assumed language \cite{fulton-thesis}.
	\end{enumerate}
When a type error is found, an intelligible error message is provided.
\item \textbf{Elaboration logic} that partially or fully compiles known regular expressions into the efficient internal representation that will be used by the regular expression matching engine (e.g., a finite automata \cite{Thompson:1968:PTR:363347.363387}) ahead of time. In most languages, this compilation step occurs at run-time, even if the pattern is fully known at compile-time, thereby introducing performance overhead into programs. If the developer is not careful to cache compiled representations, regular expressions used repeatedly in a program might be needlessly re-compiled on each use. %By performing this step ahead-of-time, these dangers can be avoided.
\item \textbf{Editor services} for quickly testing regular expression patterns against test strings,  referring to documentation or searching databases of common patterns (e.g. \cite{IntelliJRegexp}).
\end{enumerate}

%In a conventional \emph{monolithic} programming system, support for each of these features would need to be built into the language and tools. 
No system today has built-in support for all of the features enumerated above. Instead, libraries generally provide support for regular expressions by leveraging  general-purpose constructs. Unfortunately, it is impossible to fully encode the syntax and the specialized static and dynamic semantics described above in terms of general-purpose notations and abstractions like objects or inductive datatypes. Library providers have thus needed to compromise, typically by asking clients to provide regular expressions as strings, thereby deferring parsing, typechecking and elaboration to run-time, which introduces performance overhead and can lead to unanticipated run-time errors (as shown in \cite{spishak2012type}) and security vulnerabilities (due to injection attacks when user inputs are spliced into patterns, for example). Similarly, useful tools for working with regular expressions are rarely integrated into editors, and even more rarely in a way that facilitates their discovery and use directly when the developer is manipulating regular expressions. Tools that must be discovered independently and accessed externally are used less frequently  and can be more awkward than necessary \cite{Campbell:2008:DRT:1636642.1636651, Omar:2012:ACC:2337223.2337324}, leading to lower productivity \cite{Omar:2012:ACC:2337223.2337324}.

@@

\subsection{Language-Integrated Approaches}
We argue that, due to these problems, taking a language-external approach to realizing a new feature should be considered harmful and avoided whenever possible. The goal of the research being proposed here is to design \emph{language-integrated extension mechanisms} that give developers the ability to define in libraries new edit-time and compile-time features of the system that have previously been defined centrally\footnote{One might compare today's monolithic programming systems to  {centrally-planned} economies, whereas extensible\- systems more closely resemble modern market economies.} so that such language-external approaches are less frequently necessary. More specifically, we will show how control over aspects of \textbf{syntax}, \textbf{typechecking}, \textbf{elaboration} and \textbf{editor services} can be delegated to user-defined logic distributed in {libraries}, as illustrated in Figure \ref{approaches}b. 
Such libraries are called \emph{active libraries} because, rather than being passive clients of features already available in the system, they contain logic invoked by the system during development or compilation to provide new features  \cite{activelibraries}. Features implemented within active libraries can be imported as needed by the clients of libraries that rely on them, unlike features implemented by language-external means, seemingly avoiding the client compatibility problem.

>>>>

% !TEX root = omar-thesis-proposal.tex
\vspace{-25pt}
\section{Motivation}\label{motivation}
Specifying and implementing a programming language together with its supporting tools (collectively, a \emph{programming system}) that has sound theoretical foundations, helps users identify and fix errors as early as possible, supports a natural programming style, and performs well across diverse problem domains and hardware platforms remains a grand challenge in computing. Due to the increasing diversity and complexity of modern problem domains and hardware platforms, it has become clear that no small  collection of general-purpose constructs backed by a simple, all-purpose implementation can fully satisfy these criteria in all cases. Instead,  researchers and domain experts (collectively, \emph{providers}) continue to develop  specialized notations,  static and dynamic semantics, implementation strategies, optimizations, run-time systems and tools (collectively, \emph{features})  designed to address these modern challenges in different circumstances. %, researchers and domain experts must continue to design and develop, in a principled manner, specialized notations, type systems, optimizations, run-time systems and tools (collectively, \emph{features}) to support developers as they work in increasingly complex problem domains and on modern hardware systems.

@@

Ideally, a provider would develop and distribute such new features orthogonally, as libraries, so that client developers could granularly choose those that best satisfy their needs. Unfortunately this is often infeasible because libraries are vehicles for specifying a program in terms of a monolithic collection of existing features. From the perspective of a library, the language's syntax, type system and dynamic semantics are fixed in advance, the compiler and run-time system are ``black box'' implementations of this fixed language, and the other tools, like code editors and debuggers, operate according to domain-agnostic protocols also based only on a fixed language specification. As a result, providers of new system features must often take \emph{language-external approaches}, sometimes designing a new system altogether. We will argue that such approaches are problematic, and that taking them has led to an unnecessary gap between research and practice. In place of these approaches, we will advocate for \emph{language-integrated extensibility mechanisms}, and show how, by organizing new features around {types} and enforcing certain constraints, these mechanisms can be made both safe and highly expressive. We call a type that introduces new features an \emph{active type} and programming systems organized around active types \emph{actively-typed programming systems}.
%But taking a \emph{language-internal approach} to implementing a feature is the most practical. If a feature can be realized by creatively using existing language constructs and distributed as a library, clients face fewer barriers to adoption because it is easy to integrate library-based features into existing projects gradually and granularly and they leverage well-understood and well-developed mechanisms.
% But taking this approach is often \emph{not} possible today %We call designs \emph{monolithic programming systems}.

@@

\begin{enumerate}
\item \textbf{Built-in syntax for pattern literals} (e.g. \cite{perlre}) so that malformed patterns result in intelligible \emph{compile-time} parsing errors. In languages lacking such literals, run-time errors relating to pattern syntax can occur, even in well-tested, deployed code \cite{spishak2012type}. 
\item \textbf{Typechecking logic} that ensures that key constraints related to regular expressions are not violated at compile-time:
	\begin{enumerate}
	\item only appropriate values are spliced into regular expressions, to avoid splicing errors and injection attacks \cite{owasp2013}
	\item out-of-bounds backreferences are not used \cite{spishak2012type}
	\item strings known to be in the language of a regular expression are given a type that tracks this information and ensures that string manipulation operations do not inadvertently lead to a string that is not in the assumed language \cite{fulton-thesis}.
	\end{enumerate}
When a type error is found, an intelligible error message is provided.
\item \textbf{Translation logic} that partially or fully compiles known regular expressions into the efficient internal representation that will be used by the regular expression matching engine (e.g., a finite automata \cite{Thompson:1968:PTR:363347.363387}) ahead of time. In most languages, this compilation step occurs at run-time, even if the pattern is fully known at compile-time, thereby introducing performance overhead into programs. If the developer is not careful to cache compiled representations, regular expressions used repeatedly in a program might be needlessly re-compiled on each use. %By performing this step ahead-of-time, these dangers can be avoided.
\item \textbf{Editor services} for quickly testing regular expression patterns against test strings,  referring to documentation or searching databases of common patterns (e.g. \cite{IntelliJRegexp}).
\end{enumerate}

%In a conventional \emph{monolithic} programming system, support for each of these features would need to be built into the language and tools. 
No system today has built-in support for all of the features enumerated above. Instead, libraries generally provide support for regular expressions by leveraging  general-purpose constructs. Unfortunately, it is impossible to fully encode the syntax and the specialized static and dynamic semantics described above in terms of general-purpose notations and abstractions like objects or inductive datatypes. Library providers have thus needed to compromise, typically by asking clients to provide regular expressions as strings, thereby deferring parsing, typechecking and translation to run-time, which introduces performance overhead and can lead to unanticipated run-time errors (as shown in \cite{spishak2012type}) and security vulnerabilities (due to injection attacks when user inputs are spliced into patterns, for example). Similarly, useful tools for working with regular expressions are rarely integrated into editors, and even more rarely in a way that facilitates their discovery and use directly when the developer is manipulating regular expressions. Tools that must be discovered independently and accessed externally are used less frequently  and can be more awkward than necessary \cite{Campbell:2008:DRT:1636642.1636651, Omar:2012:ACC:2337223.2337324}, leading to lower productivity \cite{Omar:2012:ACC:2337223.2337324}.

@@

\subsection{Language-Integrated Approaches}
We argue that, due to these problems, taking a language-external approach to realizing a new feature should be considered harmful and avoided whenever possible. The goal of the research being proposed here is to design \emph{language-integrated extension mechanisms} that give developers the ability to define in libraries new edit-time and compile-time features of the system that have previously been defined centrally\footnote{One might compare today's monolithic programming systems to  {centrally-planned} economies, whereas extensible\- systems more closely resemble modern market economies.} so that such language-external approaches are less frequently necessary. More specifically, we will show how control over aspects of \textbf{syntax}, \textbf{typechecking}, \textbf{translation} and \textbf{editor services} can be delegated to user-defined logic distributed in {libraries}, as illustrated in Figure \ref{approaches}b. 
Such libraries are called \emph{active libraries} because, rather than being passive clients of features already available in the system, they contain logic invoked by the system during development or compilation to provide new features  \cite{activelibraries}. Features implemented within active libraries can be imported as needed by the clients of libraries that rely on them, unlike features implemented by language-external means, seemingly avoiding the client compatibility problem.

### 2014-01-16 thesis-proposal/timeline.tex
Full rough draft.

\section{Research Plan}

\subsection{Spring 2013}
I will complete the work on Ace and active type theory and submit it for publication.

\subsection{Summer 2013}
I will complete the work on active code prediction and submit it for publication.

\subsection{Fall 2013}
I will finalize all work and write the final dissertation.

>>>>

% !TEX root = omar-thesis-proposal.tex
\section{Timeline}
\begin{itemize}
\item @$\lambda$ - ICFP - Mar. 1
\item Wyvern - ECOOP or OOPSLA - late March
\item Ace - PLDI or OOPSLA - late March
\item Then write everything up.
\end{itemize}
